A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-05-12 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: more work on the rewrite

Our current understanding of probability is given in terms of the measure theoretic formalization due to Kolmogorov, and its various philosophical interpretations. However, probability in quantum mechanics, as it is widely known, is not Kolmogorovian, in the sense that it provides results that are not reproducible in standard probability spaces. It follows that our mathematical axioms for probability and their interpretations are insufficient for physical theories.

An alternative measure theoretic strategy, that has been successful is quantum mechanics, is the use of quasi-probability distributions: measures that are still unitary over the whole space, but can have negative values in small regions. However, these haven't resulted in an alternative axiomatic approach with suitable interpretations for at least, in our mind, three reasons. First, it is not clear how negative probability should be interpreted. Second, it is not clear how to characterize the space of valid quasi-probability distributions (e.g. there is no simple rule to specify which $\rho(q,p)$ are valid Wigner functions). Third, there are multiple options for quasi-probability distributions (e.g. Wigner function, Husimi distribution, Sudarshan–Glauber distribution). This lack of foundational clarity presents an additional problem: how do we know whether future physical theories will not require yet another extension?

@@

\section{Quick review of probability theory}

The standard way to characterize probability is through measure theory. Given a sample space $\Omega$, which represents the possible outcomes, and a $\sigma$-algebra $\Sigma_{\Omega}$, which represents the events, we define a probability measure $p : \Sigma_{\Omega} \to [0,1]$. Note that the probability is assigned to events, not the outcomes. This is what allows to treat discrete and continuous spaces within the same framework. In physics, all sample spaces are ultimately topological spaces, and the $\sigma$-algebra is typically the Borel algebra (i.e. the smallest $\sigma$-algebra that contains all open sets).

In classical mechanics, the sample space is phase space $M$ given by all possible values of position and momentum for all particles under considerations. Phase space is equipped with another measure, the Liouville measure $\mu : \Sigma_M \to [0, +\infty]$, that returns volume of each region. In statistical mechanics, this volume represents the count of states and is another key ingredient. The probability density over states is given by the Radon-Nikodym derivative $\rho = \frac{dp}{d\mu}$ and the entropy is given by $S(\rho) = - \int_M \rho \log \rho d\mu$. In case of a uniform distribution $\rho_U$ with support $U$, the entropy is given by $S(\rho_U) = \log \mu(U)$.\footnote{Note that entropy, the count of states and the symplectic form that characterizes phase space are deeply intertwined, yet they require different mathematical framework to express them. We suspect that they are "shadows" of a more fundamental mathematical structure, which our work on ensembles space is trying to identify.}

The above definitions do not apply to quantum mechanics. In the Hilbert space formulation, mixed states are represented by density operator (i.e. positive semi-definite self-adjoint operators with trace one). Note that if $\rho(x)$ is an integrable function such that $\int_X \rho(x) dx = 1$, the map $\psi(x) \mapsto \rho(x) \psi(x)$ is a density operator. Therefore the space of probability densities over position are mixed states in quantum mechanics. However, no joint distribution between position and momentum can be given, therefore quantum states cannot be described by the space of probability measures over a single sample space.

Alternative representations over phase space do exist, and they are widely used.(CITE) The use quasi-probability measure, meaning that the measure over the full space is still unitary (i.e. $\mu(M) =1$). However, the measure can be negative for smaller regions, which means it can also be grater than $1$, and there is no simple characterization of which quasi-probability measures are allowed and which aren't. Moreover, there is no simple link between then entropy of the distribution and the Liouville measure.

>>>>

In the past few decades, there has been a growing interest in developing approaches that generalize the notion of states and physical theories. One basic insight is that statistical/probabilistic concepts often provide a suitable basis for this generalization, and it is in this setting where most of these approaches operate. A key problem, then, is what notion of probability should be used given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics.

We reformulate the problem in the following way: given the space of all possible statistical ensembles allowed by a statistical theory, what measure theoretic tools can we use to represent each ensemble and the geometric properties of the overall space? Why exactly classical probability measures fail to capture completely quantum states? What suitable generalizations are possible, and what mathematical tools would need to be developed?

In this paper, we will characterize a generic space of ensembles as a convex bounded set of a Hausdorff locally convex topological vector space, upon which an entropy function is defined. These mathematical properties are physically grounded in the ability to experimentally define ensemble (leading to a topological space), to define statistical mixtures (leading to a convex space), to quantify the variability of the elements of the ensembles (leading to an entropy function) and to having enough real valued quantities to identify each ensemble (leading to the continuous embedding into a Hausdorff locally convex topological vector space). The question is, then, what measure theoretic tool can quantify the number of states (as in the classical Liouville measure) or the number of possible outcomes of a measurements (as in the number of orthogonal subspaces of an operator)? What measure theoretic tool can represent each ensemble? If we are to define a physically meaningful notion of probability density, both of these questions must be answered.

We will first review some results from Choquet theory, which is, to our knowledge, the most general approach to represent elements of a convex set using additive probability measures. The key insight is that, at least in the compact case, such representation is always possible, but it is not unique. Uniqueness requires the space to be a Choquet simplex, which is a generalization of the standard finite-dimensional simplex and is a particularly shaped subset of a vector space. The space of quantum ensembles is not a simplex for one key reason: it allows multiple decompositions in terms of pure states. These insights will allow to better understand exactly the limits of classical probability to describe ensembles in physical theories.

We will also review some results from the use of quasi-probability measures. While the Wigner function represents the most known instance, there are different choices. Note that the space of signed probability measures over a set $K$ is a vector space, so one can find an affine map between a suitable space of signed probability measures and the embedding vector space. Each ensemble, then, is uniquely represented with a signed probability measure. However, unlike in the Choquet case, the measure is not supported by the set of extreme points of the ensemble space. That is, there is no unique or privileged choice for this representation. Moreover, since the map is at the level of the full vector space, not just the subset with valid ensembles, there is no general way to characterize which subset of signed probability measures correspond to ensembles. Therefore, while signed measures are a powerful tool for calculation in specific problems, they cannot give us a general principled representation of the physical theory.

The above two approaches exhaust what can be done with additive measures, so we turn to the non-additive case. From the entropy, we can define the state capacity, which generalizes the notion of count of states. For each ensemble, we can define the fraction capacity, which generalizes the notion of probability. Both are non-negative sub-additive monotonic set functions, with the second being unit bounded. These measures are always unique and, since they are defined on the ensemble space itself, always have provide a physically meaningful representation. Moreover, in the compact case, these measure can be restricted on the extreme points while retaining uniqueness. (TODO: prove)

The fraction capacity is the supremum of all possible measures for the same ensemble. The classical case exactly corresponds to the simplex case, in which fraction capacity corresponds to a unique measure, and is therefore additive. The state capacity, instead, corresponds to the minimum of the entropy of all possible representations. It recovers the Liouville measure in the classical case and the dimensionality of the Hilbert space in the quantum case.

Having shown that a general approach based on non-additive measures is possible, we lay down a series of questions to explore in the future. The biggest one is whether it is possible to generalize notions such as integration, derivation and expectation to this non-additive case.  It should be noted that non-additive measures are used in the context of decision theory, so there may be a connection between the use of non-additive measures in quantum theory and interdependent complimentary choices.[TODO: find citation that sub-additive is when choices are complementary] Both the Choquet and the Sugeno integral, for example, are defined over a monotonic set-functions, so there is a chance that these tools may turn out to be useful. We see this work as starting this line of research, which will take more effort to investigate.

A final remark before we start. Too often works in the foundations of physics focus on a particular aspect without concern to a broader connection to all relevant mathematical, physical and philosophical aspects. Therefore, it is important to us that this work is a piece of a much larger puzzle, which connects cleanly and directly to  other areas, such as information geometry, functional analysis, quantum logic and so on. It is the hallmark of a true fundamental structure for physics that such connections can be easily made. Unfortunately, the nature of scientific publication requires us to publish a narrow slice of the work to fit the scope of a particular journal. Ironically, this means that the most important motivations for our approach are out of scope of this article, as we will concentrate only to the connections to measure theory.

\section{Ensemble space}

Conceptually, a statistical ensemble is a collection of outputs, taken at once, of a reproducible preparation procedure for a physical system. In the classical case, this corresponds to a probability distribution over a discrete set of outcomes (i.e. a fair coin) or over phase space (i.e. the symplectic manifold representing position and momentum for a classical particle). In the quantum case, this corresponds to a mixed state (i.e. a positive semi-definite self-adjoint operator of trace one). An ensemble space is the collection of all possible ensembles defined in a physical theory, together with a basic mathematical structure that all physical theories must possess. The development of a full theory of ensembles goes beyond the scope of this work, and we refer to the appendix and our ongoing work (cite).

For the purpose of this paper, we will assume an ensemble space $\Ens$ is a convex subset of a Hausdorff locally convex topological vector space $V$. The topology is the physically natural topology, where each open set represents a verifiable statement. The convex structure represents statistical mixing. For example, $p \ens[a] + \bar{p} \ens[b]$ represents the statistical mixture of two ensembles $\ens[a]$ and $\ens[b]$ with probability $p$ and $\bar{p} = 1 - p$. Only finite mixtures are guaranteed, while the topology will decide whether an infinite mixture $\sum_{i}^{\infty} p_i \ens_i$ converges in the space or not. We also assume that there is a countable family of affine continuous functions $F_i : \Ens \to \mathbb{R}$ that represent statistical quantities and fully identify each ensemble. That is, $F_i(p \ens[a] + \bar{p} \ens[b]) = p F_i(\ens[a]) + \bar{p} F_i(\ens[b])$ and $\ens[a] = \ens[b]$ if and only if $F_i(\ens[a]) = F_i(\ens[b])$ for all $i$.\footnote{It is an open question whether all ensemble spaces must possess this property.} For example, we can imagine $F_i$ representing the expectation of all polynomials of position and momentum.\footnote{This requires all probability distributions to converge to zero at infinity faster than any polynomial.} Mathematically, these statistical quantities can be used to define semi-norms on the space, giving us a locally convex topology.

Additional, each ensemble space $\Ens$ is equipped with a continuous entropy function $S : \Ens \to \mathbb{R}$. The entropy of a statistical mixture satisfies the following bounds
\begin{equation}
	\begin{aligned}
		p S(\ens[a]) + (1-p) &S(\ens[b]) \leq S(p \ens[a] + (1-p) \ens[b]) \\
		&\leq I(p,\bar{p}) + p S(\ens[a]) + (1-p) S(\ens[b])
	\end{aligned}
\end{equation}
where $I(p,\bar{p}) = - p \log p - \bar{p} \log(\bar{p})$.\footnote{In the larger work, we can show that the Shannon entropy is the only upper bound possible.} The bounds come from assuming that the entropy quantifies the variability of each ensemble. Intuitively, the variability cannot decrease during mixing, and will have an upper bound when mixing two ensembles that are non-overlapping, or orthogonal. In fact, we define two ensembles $\ens[a] \ortho \ens[b]$ to be orthogonal when they mixture gives a maximal entropy increase. This does recover the notion of orthogonality in the corresponding classical and quantum vector spaces of statistical ensembles. Lastly, mixtures preserve orthogonality in the sense that mixing non-overlapping ensembles still gives non-overlapping ensembles. More precisely, $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$.

To summarize, the above mathematical structure represents the requirements that ensembles must be experimentally well-defined, allow statistical mixture, can be identified with statistical quantities and provide a well-defined entropy.\footnote{While the full physical justification of the above mathematical structure is beyond the scope of this work, it is important to us that such justification exists and this physically straight forward. This is the essence of what we call Physical Mathematics. We will summarize some of these justifications in the appendix, leaving the full and more updated discussion in our work in progress.} We now turn to the question of how we can represent both the entropy and each ensemble in terms of some measure theoretic structure.

\section{Additive measure and Choquet theory}

Since ensemble spaces are a convex subset of a locally convex topological vector space, they can be connected to Choquet theory. First of all, let $E$ be a convex subset of a locally convex space $V$. We say that $\ens \in E$ is represented by a probability measure $\mu : \Sigma_{E} \to [0,1]$ if $f(\ens) = \int_X f d\mu$ for every continuous linear functional $f$ on $V$. The Choquet theorem states:
\begin{thrm}[Choquet]
	Let $E$ be a metrizable compact convex subset of a locally convex space $V$, and let $\ens \in E$. Then there exists a probability measure $\mu_{\ens}$ on $X$ which represents $\ens$ and is supported by the extreme points of $X$.
\end{thrm}

First, let us understand the physical significance. An ensemble $\ens$ can be seen as a probability measure $\mu$ over the ``pure states'' (i.e. the extreme points of $E$) if the value for every statistical quantity $f$ for $\ens$ can be recovered as an expectation over the pure states using the probability measure. That is, all statistical properties of $\ens$ can be recovered by the probability measure. This is what allows to think go $\ens$ as a probability distribution. Note, however, that the Choquet theorem does not guarantee uniqueness. The probability measure is unique if and only if $E$ is a Choquet simplex, which is a generalization of the finite dimensional simplex. The characterization of the Choquet simplex is fairly technical, so we will simply define the Choque simplex by the property of having a single decomposition in terms of extreme points.

We should now look at the requirements of the theorem, to see whether we can always apply it to an ensemble space. First of all, and ensemble space $\Ens$ is a convex subset of a convex sapce $V$. Moreover, since every Hausdorff first-countable topological vector space is metrizable, $\Ens$ is metrizable. However, $\Ens$ is not necessarily compact. In fact, $\Ens$ is not necessarily a closed subset of $V$. Consider the space of probability measures over $\mathbb{R}$. Since $\mathbb{R}$ is not compact, the space of probability measure over $\mathbb{R}$ cannot be compact.\footnote{TODO: Check with mathematicians} Moreover, the set of extreme points is given by the Dirac measures (i.e. probability one over a single point), which would have minus infinite entropy, and therefore cannot be part of the ensemble space. Therefore the theorem does not apply in general.

We can apply Choquet theorem in the case where the closure of $\Ens$ in $V$ is a compact set $E$, which will also be convex\footnote{The closure of a convex set is convex} and metrizable. In this case, we are guaranteed that all ensembles allows a representation through a probability on points that may corresponds idealized ensembles (i.e. the Dirac measure) that are not, even conceptually, realizable. Many interesting cases fall into this restriction: the ensemble spaces of classical probability distribution over finitely many discrete cases fall in this restriction; the subset of ensembles over finite ranges of position and momentum; the ensemble spaces of all finite dimensional quantum systems.

Note that ensembles over classical spaces will be represented by a unique probability measure, therefore the corresponding closure of the ensemble space is a Choquet simplex. However, quantum mixed states can be decomposed in multiple ways in terms of pure states, and therefore the closure of a quantum ensemble space is not a simplex.

% space is metrizable https://proofwiki.org/wiki/Birkhoff-Kakutani_Theorem/Topological_Vector_Space

Chouqet theory tells, in the realm where it can be applied, that classical probability fails in quantum mechanics precisely because the ensemble space of a quantum system allows multiple decomposition.

\section{Quasiprobability}

Another way to represent quantum ensembles is through the use of quasiprobability distributions. Given a space $X$, a measure $\mu$ is affine if $\mu(X)$. All probability measures are affine measures, but the reverse is not true as affine measures can be negative on particular regions. The Wigner function, for example, can be negative over small regions of phase space.

Given a space $X$, the space of all affine measure is a vector space, not just a convex subset as in the case of the space of probability measures. This means that, given an ensemble space, we may find a space $X$ such that the space of affine measures over $X$ is isomorphic to the embedding vector space $V$ as topological vector spaces. In this case, any element $v\in V$ can be represented uniquely by an affine measure $\mu_{v}$, regardless of whether they are actual ensembles or not.\footnote{The question of whether such an $X$ always exists is left open. We suspect a link exists between the statistical quantities $F_i$ that fully identify each ensemble and the space $X$.} Because the statistical quantities $F_i$ are affine functions, we will have $F_i(v) = \int_X F_i d\mu_{v}$.

In general, since the ensemble space is bounded along every direction, not all affine measure will correspond to an ensemble space. Whether all probability measure correspond to a valid ensemble, or whether affine measures are even used, correspond to relationships between $\Ens$ and the space of probability measures $M_p(X)$. If $\Ens \subset M_p(X)$, then all states are represented by a probability measure. However, this is not a ``probability'' in the standard sense of the word. For example, the Husimi Q representation is always non-negative at every point, but two points to do not represent mutually exclusive events as they are not orthogonal (i.e. their inner product is non-zero and they do not maximize the entropy increase during mixture). Mathematically, the issue is that the points $X$ of the Husimi Q representation are not the extreme points of the closure ensembles space $\overline{\Ens}$. In general, $\Ens$ and $M_p(X)$ will overlap, but neither will contain the other. For example, in the Wigner representation the only pure states (i.e. the extreme points of $\Ens$) represented by a probability measure are the Gaussian wave-packets. This means that any other non-negative Wigner function will be in the convex hull of Gaussian wave-packets, and therefore will allow a decomposition in terms of them.

Note that if $\overline{\Ens} = M_p(X)$, then the space of the extreme point is exactly the space of Dirac measures, and therefore $X$ corresponds to the pure states. Alternatively, either one ensemble is represented by an affine measure, or not all probability measures can be understood as mixtures of extreme points. If $\overline{\Ens}$ is compact, then this is exactly when $\overline{\Ens}$ is not a simplex and multiple decompositions are allowed.

This alternative way of looking at measure theoretic representations, then, is consistent with the previous one. While quasiprobability distributions are very useful in practice, we want to stress that there is no single choice for $X$. To solve a specific problem, this may actually be useful as one can choose the most convenient $X$ for the problem. To develop a general theory for physics, however, this is not useful as each representation is not general enough.

\section{Old introduction}

@@
